Founder allele effects for hotspot traits

### 2019-09-07 analysis/paper/Rmd/methods.Rmd
removed heatmaply dependency; added iheatmapr & pheatmap
updated methods

@keller2018genetic performed univariate QTL mapping for all expression traits. They
defined an expression trait hotspot to be a 4-Mb region that affects at least 100 nonlocal 
expression traits.
They identified five expression trait hotspots in their
study of pancreatic islet gene expression in 378 Diversity Outbred mice.

In analyzing the data from @keller2018genetic, we first established criteria for identifying local and nonlocal transcripts at each of the five hotspots. We focused on those local 
traits that strongly associated with QTL within the hotspot. We set the univariate LOD threshold for 
local traits at 15. We limited consideration to those local traits for which the middle of the gene lay 
within the hotspot interval. 
All nonlocal traits demonstrated univariate LOD peaks that met
the genome-wide significance threshold, 7.18 [@keller2018genetic]. The nonlocal expression traits arose 
from genes
on other chromosomes, while the local traits reflected expression levels for
genes on the same chromosome as the hotspot. We ignored transcripts for genes on the hotspot chromosome 
but outside the hotspot interval.

For each hotspot, we performed both pairwise pleiotropy tests and linear regression-based 
mediation analyses. For every pair of traits at a hotspot, we calculated a pleiotropy test 
statistic[@boehm2019testing]. Due to the computational burden, we didn't determine bootstrap p-values for these 
tests. For every local trait - nonlocal trait pair, we also performed a linear 
regression-based mediation analysis [@chick2016defining,@keller2018genetic]. We describe in detail both procedures below.

@@

For each hotspot, we tested all pairs of traits, regardless of whether the traits were local or nonlocal. 
We performed a
two-dimensional, two-QTL scan over a genomic region that included the entirety of the
hotspot. We fitted bivariate linear mixed effects models (Equations \@ref(eq:mvlmm), \@ref(eq:g), and \@ref(eq:e)) at each ordered pair of markers in a two-dimensional grid [@boehm2019testing].

@@

From the two-dimensional scans, we calculated pleiotropy likelihood ratio test (LOD) statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

@@

In all of our mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship. 
We only considered as potential mediators those local traits with strong univariate LODs.
In Equations~\ref{eq:model1} through \ref{eq:model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator (a local expression trait, in our case), $\beta$ is the effect
of the mediator on the complex trait (after adjusting for other covariates in the model), 1 is a vector with all entries set to 1, $\alpha$
is a parameter that corresponds to the mean trait value (after adjusting for other covariates in the model), and $E$ is a vector of normally distributed random errors.

>>>>

@keller2018genetic performed univariate QTL mapping for all expression traits. They
defined an expression trait hotspot to be a 4-Mb region that affects at least 100 nonlocal 
expression traits.
They identified five expression trait hotspots - on Chromosomes 2, 5, 7, 11, and 13 - in their data.

In analyzing the data from @keller2018genetic, we first established criteria for identifying local and 
nonlocal transcripts at each of the five hotspots. We defined a "local trait" to be one whose corresponding gene lay within the hotspot. We focused on those local 
traits that strongly associated with QTL within the hotspot. To limit the scope of our analyses, 
we only studied those expression traits with univariate LOD greater than 15.

We defined a "nonlocal trait" to be one with univariate evidence of association (LOD at least 7.18, the genome-wide threshold)
with a QTL in the hotspot. Additionally, we required that the nonlocal trait's gene not lie on the hotspot's chromosome. We ignored transcripts for genes on the hotspot chromosome 
but outside the hotspot interval.

For each hotspot, we performed both pairwise pleiotropy tests and linear regression-based 
mediation analyses. For every pair of traits at a hotspot, we calculated a pleiotropy test 
statistic[@boehm2019testing]. Due to the computational burden, we didn't determine bootstrap p-values for these 
tests. For every local trait - nonlocal trait pair, we also performed a linear 
regression-based mediation analysis [@chick2016defining,@keller2018genetic]. We describe both procedures below.

@@

For each hotspot, we tested all pairs of traits, regardless of whether the traits were local or nonlocal. For each test, we performed a
two-dimensional, two-QTL scan over a genomic region that included the entirety of the
hotspot. We fitted bivariate linear mixed effects models (Equations \@ref(eq:mvlmm), \@ref(eq:g), and \@ref(eq:e)) at each ordered pair of markers in a two-dimensional grid [@boehm2019testing].

@@

From the two-dimensional scan results, we calculated pleiotropy likelihood ratio test (LOD) statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

@@

In all of our mediation analyses, the nonlocal traits are the targets (Y), while the local
transcripts serve as candidate mediators (M) of the QTL-target relationship. 
We only considered as potential mediators those local traits with strong univariate LODs.
In Equations~\ref{eq:model1} through \ref{eq:model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator (a local expression trait, in our case), $\beta$ is the effect
of the mediator on the complex trait (after adjusting for other covariates in the model), 1 is a vector with all entries set to 1, $\alpha$
is a parameter that corresponds to the mean trait value (after adjusting for other covariates in the model), and $E$ is a vector of normally distributed random errors.

### 2019-09-07 analysis/paper/preliminary/allele-effects-exploration.Rmd
updated methods

```{r define}
library(qtl2hotspots)

@@

```{r, define-submap}
figdir <- here::here("analysis", "figures", "Chr5-allele-effects-plots")
names(s1coef_out) -> ids
# find indices on Chr5 for boundaries of hotspot
which(map$`5` > hot_lower)[1] -> start_index
which(map$`5` > hot_upper)[1] -> stop_index
submap <- list()
submap$`5` <- map$`5`[start_index:stop_index]
mylist <- list()
for (i in 1:203){
  mylist[[i]] <- s1out[ , i, drop = FALSE]
}
```

@@

```{r blup-plots}
purrr::pmap(.l = list(s1coef_blup, as.list(ids), mylist), 
           .f = function(x, y, z){
             main <- local_nonlocal_annot %>%
               ungroup() %>%
               filter(annot.id == y) %>%
               select(annot.id, symbol) %>%
               paste0()
             qtl2::plot_coefCC(x, map = submap, main = main, scan1_output = z)
           }
)
```

@@

## Bar plots for blup effects

```{r, barplots-blup, fig.height = 20, fig.width = 8}
foobar <- local_nonlocal_annot %>%
  ungroup() %>%
  (function(x) x[ord2, ]) %>%
  select(symbol, 26:33) %>%
  tidyr::gather(paste0("b", LETTERS[1:8]), key = "founder_allele", value = "blup_effect")
foobar2 <- arrange(transform(foobar,
             symbol=factor(symbol,levels=rn)),symbol)
foobar2 %>% 
  ggplot() + geom_col(aes(y = effect, x = founder_allele, fill = founder_allele)) +
  ylim(c(-1.5, 1.5)) + 
  facet_wrap(symbol ~ ., ncol = 4) #+ theme(strip.text.x = element_blank())
ggsave(here::here("analysis", "figures", "allele_effects_at_peaks.png"), limitsize = FALSE)
```

@@

```{r BC-v-FG}
local_nonlocal_annot %>%
  ungroup() %>%
  (function(x) x[ord2, ]) %>%
  select(symbol, 18:25) %>%
    rename("A" = `1`, "B" = `2`, "C" = `3`, "D" = `4`, "E" = `5`, "F" = `6`, G = `7`, H = `8`
         ) %>%
  mutate(contrast_BC_FG = abs((B + C) - (F + G)), row_num = 1:203) %>%
  ggplot() + geom_point(aes(y = contrast_BC_FG, x = row_num))
```

```{r BC-v-G}
p <- local_nonlocal_annot %>%
  ungroup() %>%
  (function(x) x[ord2, ]) %>%
  select(symbol, 18:25) %>%
    rename("A" = `1`, "B" = `2`, "C" = `3`, "D" = `4`, "E" = `5`, "F" = `6`, G = `7`, H = `8`
         ) %>%
  mutate(contrast = abs((B + C) / 2 - G), row_num = 1:203) %>%
  ggplot() + geom_point(aes(y = contrast, x = row_num)) + geom_smooth(aes(y = contrast, x = row_num))
plotly::ggplotly(p)
```

@@

```{r B:G}
p <- local_nonlocal_annot %>%
  ungroup() %>%
  select(symbol, cA:cH, local_tx) %>%
  (function(x) x[ord2, ]) %>%
  mutate(contrast = abs(cB - cG), row_num = 1:203) %>%
  ggplot() + geom_point(aes(y = contrast, x = row_num, colour = local_tx, symbol = symbol)) + geom_smooth(aes(y = contrast, x = row_num))
plotly::ggplotly(p)
```

@@

>>>>

```{r define-ult}
library(qtl2hotspots)

@@

```{r, define-submap}
figdir <- here::here("analysis", "figures", "Chr5-allele-effects-plots")
names(s1coef_out) -> ids # need to run inline code above to define s1coef_out
# find indices on Chr5 for boundaries of hotspot
which(map$`5` > hot_lower)[1] -> start_index
which(map$`5` > hot_upper)[1] -> stop_index
submap <- list()
submap$`5` <- map$`5`[start_index:stop_index]
mylist <- list()
for (i in 1:203){
  mylist[[i]] <- s1out[ , i, drop = FALSE]
}
```

@@

@@

@@

```{r BC-v-FG}
local_nonlocal_annot %>%
  ungroup() %>%
  (function(x) x[ord2, ]) %>%
  dplyr::select(symbol, 18:25) %>%
  mutate(contrast_BC_FG = abs((cB + cC) - (cF + cG)), row_num = 1:203) %>%
  ggplot() + geom_point(aes(y = contrast_BC_FG, x = row_num))
```

```{r BC-v-G}
p <- local_nonlocal_annot %>%
  ungroup() %>%
  (function(x) x[ord2, ]) %>%
  dplyr::select(symbol, 18:25) %>%
  mutate(contrast = abs((cB + cC) / 2 - cG), row_num = 1:203) %>%
  ggplot() + geom_point(aes(y = contrast, x = row_num)) + geom_smooth(aes(y = contrast, x = row_num))
plotly::ggplotly(p)
```

@@

```{r B:G2}
p <- local_nonlocal_annot %>%
  ungroup() %>%
  select(symbol, cA:cH, local_tx) %>%
  (function(x) x[ord2, ]) %>%
  mutate(contrast = abs(cB - cG), row_num = 1:203) %>%
  ggplot() + geom_point(aes(y = contrast, x = row_num, colour = local_tx, symbol = symbol)) + geom_smooth(aes(y = contrast, x = row_num))
plotly::ggplotly(p)
```

@@

## Calculate all partitions & contrasts

```{r}
pp <- expand.grid(1:3, 
                  1:3,
                  1:3, 
                  1:3,
                  1:3, 
                  1:3,
                  1:3, 
                  1:3
                  )
effects_mat <- local_nonlocal_annot %>%
  ungroup() %>%
  select(cA:cH) %>%
  as.matrix()

```{r}
pp_tib <- pp %>%
  as_tibble()
effects_tib <- effects_mat %>%
  as_tibble()

ef <- pp_tib %>%
  apply(MARGIN = 1, 
        FUN = function(x){apply(X = effects_tib, 
                                MARGIN = 1, 
                                FUN = function(a, 
                                               b = x){
                                  qtl2hotspots::calc_contrast(grouping = b, 
                                                              effects = a)
                                  }
                                )}
        )

ef2 <- ef %>%  apply(FUN = function(x) x[!is.na(x)], MARGIN = 1) %>% 
  lapply(FUN = function(x) bind_rows(x)) %>%
  purrr::map2(local_nonlocal_annot$symbol, .f = function(x,y){x %>% mutate(symbol = y)}) %>%
  bind_rows() 
ef2 %>%
  group_by(contrast) %>%
  summarise(v = var(value)) %>%
  arrange(desc(v))
```

```{r}
rn_tib <- tibble::tibble(symbol = rn, row_number = 1:203)
ef2 %>%
  filter(contrast == "B_F") %>%
  inner_join(rn_tib) %>%
  inner_join(local_nonlocal_annot, by = "symbol") %>%
  filter(!local_tx) %>%
  ggplot() + geom_point(aes(x = row_number, y = value)) + geom_vline(aes(xintercept = 68.5))
```

## Defining the boundaries from the heatmap

I need to examine the heatmap to see how many diagonal "blocks" there are, and where the endpoints are, from 1:203, for every block.

First, note that 1:182 are all nonlocal traits. Now, Adamts15 through Cntnap2, inclusive, seems to define the first block.

The second block starts right above Cntnap2, ie, with Elmo1 and ends roughtly at Cep68.

Let's look at the first "block".

```{r}
which(rn_tib$symbol == "Elmo1")
which(rn_tib$symbol == "Adamts15")
abs_diff <- function(x){abs(x[1] - x[2])}
notgroup1 <- rn_tib %>%
  mutate(group1_member = row_number <= 68) %>%
  inner_join(ef2) %>%
  group_by(contrast, group1_member) %>%
  summarise(variance = var(value), mean = mean(value)) %>%
  arrange(contrast) %>%
  ungroup() %>%
  group_by(contrast) %>%
  filter(!group1_member) %>%
  arrange(variance) %>% 
  rename(variance_notgroup1 = variance, mean_notgroup1 = mean)

rn_tib %>%
  mutate(group1_member = row_number <= 68) %>%
  inner_join(ef2) %>%
  group_by(contrast, group1_member) %>%
  summarise(variance = var(value), mean = mean(value)) %>%
  arrange(contrast) %>%
  ungroup() %>%
  group_by(contrast) %>%
  filter(group1_member) %>%
  arrange(variance) %>% 
  rename(variance_group1 = variance, mean_group1 = mean) %>%
  inner_join(notgroup1, by = "contrast") %>%
  dplyr::select(- group1_member.x, - group1_member.y)

```{r}
ef2 %>%
  filter(contrast == "B_G") %>%
  inner_join(rn_tib) %>%
  inner_join(local_nonlocal_annot, by = "symbol") %>%
  filter(!local_tx) %>%
  ggplot() + geom_point(aes(x = row_number, y = value)) + geom_vline(aes(xintercept = 68.5)) + geom_smooth(aes(x = row_number, y = value))

```{r}
rn_tib %>%
  mutate(group1_member = row_number <= 68) %>%
  inner_join(ef2) %>%
  group_by(contrast) %>%
  summarise(t_stat = t.test(value ~ group1_member)$statistic) %>%
  arrange(desc(t_stat))
```

```{r}
ef2 %>%
  filter(contrast == "AGH_CEF") %>%
  inner_join(rn_tib) %>%
  inner_join(local_nonlocal_annot, by = "symbol") %>%
  filter(!local_tx) %>%
  ggplot() + geom_point(aes(x = row_number, y = value)) + geom_vline(aes(xintercept = 68.5))

```{r, define-contrasts-5-1}
contr <- ef2 %>%
  filter(contrast == "AGH_CEF") %>%
  inner_join(rn_tib) %>%
  arrange(row_number)
```

```{r, order-annotations-5-1}
# all annots 
annots_ord <- list(lod_diffs_tibble, 
                   lod_diff_proportions_tibble, 
                   allele_effects_tibble,
                   peak_positions, 
                   local_tibble,
                   lod_tibble, 
                   contr
                    ) %>% 
  purrr::map(.f = function(df, rn){
    out <- df[match(x = rn, table = df$symbol), ]
    return(out)
  }, 
  rn = rn)
ann7 <- annots_ord[[7]] %>% # contrast
  dplyr::select(value) %>%
  as.matrix() %>%
  (function(x){rownames(x) <- annots_ord[[7]]$symbol; return(x)})
```

```{r}
iheatmapr::main_heatmap(mymat, show_colorbar = FALSE, layout = list(font = list(size = 4))) %>%
  iheatmapr::add_col_labels() %>%
  iheatmapr::add_col_annotation(ann5, show_colorbar = FALSE) %>%
  iheatmapr::add_main_heatmap(ann3, name = "signal3", size = 0.2, show_colorbar = FALSE) %>%
    iheatmapr::add_col_labels() %>%

  iheatmapr::add_main_heatmap(ann2, name = "signal2", size = 0.5, show_colorbar = FALSE) %>%
  iheatmapr::add_col_labels() %>%
  iheatmapr::add_main_heatmap(ann1, name = "signal1", size = 0.5, show_colorbar = FALSE) %>%
  iheatmapr::add_row_annotation(ann5, side = "left", show_colorbar = TRUE) %>%
  iheatmapr::add_col_labels() %>%
  iheatmapr::add_row_annotation(ann4, side = "left", show_colorbar = TRUE) %>%
  iheatmapr::add_row_annotation(ann6, side = "left", show_colorbar = TRUE) %>%
  iheatmapr::add_row_annotation(ann7, side = "left", show_colorbar = TRUE) %>%
  iheatmapr::add_row_labels()

# Other ideas

What results do we get when we try to "find the contrast" for other hotspots: 2, 7, 11, 13?
